<<<<<<< HEAD
---
title: "Predicting Type 2 Diabetes: A Comparative Machine Learning Analysis"
author: "Data Science Capstone Project"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Table of Contents"
    number-sections: true
    theme: cosmo
    code-fold: true
    code-tools: true
    fig-width: 9
    fig-height: 6
    fig-align: center
    embed-resources: true
    smooth-scroll: true
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

Type 2 diabetes is a major global health burden, affecting over 500 million people worldwide. Early identification of at-risk individuals allows for timely lifestyle interventions that can delay or prevent onset. This analysis uses the **Pima Indians Diabetes Dataset** — 768 health measurements from Pima Indian women aged 21 and older — to identify the most influential clinical predictors of diabetes and compare three classification models: Logistic Regression, LASSO Regression, and Random Forest.

**Research Questions:**

-   Which clinical variables are most influential in predicting type 2 diabetes?
-   Which model best balances predictive accuracy and interpretability for this clinical task?

------------------------------------------------------------------------

# Setup

```{r packages}
rm(list = ls())
set.seed(123)

required_packages <- c(
  "tidyverse", "mlbench", "caret", "pROC",
  "glmnet", "mice", "gridExtra", "reshape2",
  "randomForest", "knitr", "kableExtra", "scales", "viridis"
)

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}

# Color palette
palette_neg <- "#2E86AB"   # Blue  – No Diabetes
palette_pos <- "#A23B72"   # Purple – Diabetes
palette_neu <- "#F18F01"   # Orange – neutral / third element
```

------------------------------------------------------------------------

# Data & Preprocessing

We load the `PimaIndiansDiabetes2` dataset, which includes medically meaningful predictors and a binary outcome indicating diabetes diagnosis within five years. Several variables contain missing values (notably insulin at 49% and triceps at 30%), which are addressed using **Multiple Imputation by Chained Equations (MICE)** with predictive mean matching rather than simple deletion or zero-substitution.

```{r load-and-impute}
data("PimaIndiansDiabetes2", package = "mlbench")
diabetes_raw <- PimaIndiansDiabetes2

# Impute missing values
imp <- mice(diabetes_raw[, -9], m = 5, method = "pmm",
            seed = 123, printFlag = FALSE)
diabetes <- complete(imp, 1)
diabetes$diabetes <- diabetes_raw$diabetes

continuous_vars <- c("pregnant","glucose","pressure","triceps",
                     "insulin","mass","pedigree","age")
```

**Train / Test Split** — 70% training, 30% testing, stratified by outcome.

```{r split}
idx        <- createDataPartition(diabetes$diabetes, p = 0.70, list = FALSE)
train_data <- diabetes[idx, ]
test_data  <- diabetes[-idx, ]
```

------------------------------------------------------------------------

# Exploratory Data Analysis

## Predictor Distributions by Outcome

The density plots below reveal which variables show the clearest separation between diabetic and non-diabetic patients. Glucose, BMI, and age stand out as the most discriminating predictors.

```{r ridge-plot, fig.height=8}
diabetes |>
  select(all_of(continuous_vars), diabetes) |>
  pivot_longer(-diabetes, names_to = "Variable", values_to = "Value") |>
  mutate(Outcome = ifelse(diabetes == "neg", "No Diabetes", "Diabetes")) |>
  ggplot(aes(x = Value, fill = Outcome)) +
  geom_density(alpha = 0.65) +
  scale_fill_manual(values = c("No Diabetes" = palette_neg, "Diabetes" = palette_pos)) +
  facet_wrap(~Variable, scales = "free", ncol = 2) +
  labs(title = "Predictor Distributions by Diabetes Status",
       x = NULL, y = "Density", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "top",
        strip.text = element_text(face = "bold"))
```

## Box Plots: Top Predictors vs. Outcome

```{r boxplots, fig.height=5}
top_vars   <- c("glucose", "mass", "age", "pedigree")
top_labels <- c("Glucose (mg/dL)", "BMI (kg/m2)", "Age (years)", "Pedigree Function")

bp_list <- mapply(function(v, lab) {
  ggplot(diabetes, aes_string(x = "diabetes", y = v, fill = "diabetes")) +
    geom_boxplot(notch = TRUE, outlier.alpha = 0.4, show.legend = FALSE) +
    scale_x_discrete(labels = c("neg" = "No Diabetes", "pos" = "Diabetes")) +
    scale_fill_manual(values = c("neg" = palette_neg, "pos" = palette_pos)) +
    labs(title = lab, x = NULL, y = NULL) +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 10))
}, top_vars, top_labels, SIMPLIFY = FALSE)

grid.arrange(grobs = bp_list, ncol = 4,
             top = "Key Predictors by Diabetes Status")
```

## Correlation Heatmap

Understanding predictor correlations is important for interpreting LASSO's variable selection behavior and for detecting multicollinearity in logistic regression.

```{r heatmap, fig.height=6, fig.width=6}
cor_melt <- diabetes |>
  select(all_of(continuous_vars)) |>
  cor() |>
  melt()

ggplot(cor_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(value, 2)), size = 3, fontface = "bold") +
  scale_fill_gradient2(low = palette_neg, mid = "white", high = palette_pos,
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  labs(title = "Predictor Correlation Matrix", x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        axis.text.y = element_text(face = "bold"),
        plot.title  = element_text(face = "bold")) +
  coord_fixed()
```

## Summary Statistics

```{r summary-table}
diabetes |>
  group_by(diabetes) |>
  summarise(across(all_of(continuous_vars),
                   ~paste0(round(mean(.), 1), " (+-", round(sd(.), 1), ")")),
            .groups = "drop") |>
  mutate(diabetes = ifelse(diabetes == "neg", "No Diabetes", "Diabetes")) |>
  pivot_longer(-diabetes, names_to = "Variable") |>
  pivot_wider(names_from = diabetes, values_from = value) |>
  kbl(caption = "Mean (+-SD) by Diabetes Status",
      col.names = c("Variable", "No Diabetes", "Diabetes")) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

------------------------------------------------------------------------

# Model Building

## Logistic Regression

Logistic regression serves as our interpretable baseline. We apply bidirectional stepwise selection (AIC) to retain only the most informative predictors.

```{r logistic}
log_full <- glm(diabetes ~ ., data = train_data, family = binomial)
log_step <- step(log_full, direction = "both", trace = 0)

# Predictions
log_prob_tr  <- predict(log_step, train_data, type = "response")
log_prob_te  <- predict(log_step, test_data,  type = "response")
log_class_tr <- factor(ifelse(log_prob_tr > 0.5, "pos", "neg"), levels = c("neg","pos"))
log_class_te <- factor(ifelse(log_prob_te > 0.5, "pos", "neg"), levels = c("neg","pos"))

log_cm_tr  <- confusionMatrix(log_class_tr, train_data$diabetes, positive = "pos")
log_cm_te  <- confusionMatrix(log_class_te, test_data$diabetes,  positive = "pos")
log_roc_tr <- roc(train_data$diabetes, log_prob_tr, quiet = TRUE)
log_roc_te <- roc(test_data$diabetes,  log_prob_te, quiet = TRUE)
```

### Odds Ratios

Variables with OR \> 1 and a confidence interval that does not cross 1 are statistically significant risk factors.

```{r odds-ratios}
or_df <- data.frame(
  Variable = names(coef(log_step)),
  OR       = exp(coef(log_step)),
  CI_Lo    = exp(confint.default(log_step)[,1]),
  CI_Hi    = exp(confint.default(log_step)[,2])
) |> filter(Variable != "(Intercept)") |>
  mutate(across(where(is.numeric), ~round(., 3)))

kbl(or_df, caption = "Logistic Regression: Odds Ratios with 95% CI",
    col.names = c("Variable","OR","Lower CI","Upper CI")) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) |>
  row_spec(which(or_df$OR > 1 & or_df$CI_Lo > 1), bold = TRUE,
           color = "white", background = palette_pos)

ggplot(or_df, aes(x = reorder(Variable, OR), y = OR,
                  ymin = CI_Lo, ymax = CI_Hi, color = OR > 1)) +
  geom_pointrange(size = 0.9, linewidth = 1) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("FALSE" = palette_neg, "TRUE" = palette_pos),
                     labels = c("No/Neutral Effect", "Increased Risk"), name = NULL) +
  coord_flip() +
  labs(title    = "Logistic Regression: Odds Ratios with 95% CI",
       subtitle = "Purple = statistically significant risk factor (OR > 1, CI excludes 1)",
       x = NULL, y = "Odds Ratio") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), legend.position = "top")
```

------------------------------------------------------------------------

## LASSO Regression

LASSO applies an L1 penalty that shrinks weaker coefficients to zero, producing an automatically selected subset of predictors. The optimal penalty (lambda) is chosen via 10-fold cross-validation.

```{r lasso}
x_tr <- model.matrix(diabetes ~ ., train_data)[,-1]
y_tr <- ifelse(train_data$diabetes == "pos", 1, 0)
x_te <- model.matrix(diabetes ~ ., test_data)[,-1]

cv_lasso    <- cv.glmnet(x_tr, y_tr, alpha = 1, family = "binomial", nfolds = 10)
lasso_model <- glmnet(x_tr, y_tr, alpha = 1, family = "binomial",
                      lambda = cv_lasso$lambda.min)

# Predictions
las_prob_tr  <- predict(lasso_model, x_tr, type = "response")[,1]
las_prob_te  <- predict(lasso_model, x_te, type = "response")[,1]
las_class_tr <- factor(ifelse(las_prob_tr > 0.5, "pos", "neg"), levels = c("neg","pos"))
las_class_te <- factor(ifelse(las_prob_te > 0.5, "pos", "neg"), levels = c("neg","pos"))

las_cm_tr  <- confusionMatrix(las_class_tr, train_data$diabetes, positive = "pos")
las_cm_te  <- confusionMatrix(las_class_te, test_data$diabetes,  positive = "pos")
las_roc_tr <- roc(train_data$diabetes, las_prob_tr, quiet = TRUE)
las_roc_te <- roc(test_data$diabetes,  las_prob_te, quiet = TRUE)
```

### LASSO: Selected Coefficients

```{r lasso-coef-plot}
las_coef_df <- data.frame(
  Variable = rownames(as.matrix(coef(lasso_model))),
  Coef     = as.matrix(coef(lasso_model))[,1]
) |> filter(Variable != "(Intercept)", Coef != 0) |>
  arrange(desc(abs(Coef)))

ggplot(las_coef_df, aes(x = reorder(Variable, abs(Coef)), y = Coef, fill = Coef > 0)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("FALSE" = palette_neg, "TRUE" = palette_pos)) +
  coord_flip() +
  labs(title    = "LASSO: Non-Zero Coefficients After Penalization",
       subtitle = "Purple = increases risk; Blue = decreases risk; absent = shrunk to zero",
       x = NULL, y = "Coefficient") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

------------------------------------------------------------------------

## Random Forest

Random Forest is an ensemble of 500 decision trees that reduces variance through bagging and random feature subsampling. It naturally captures non-linear relationships and feature interactions that linear models may miss.

```{r random-forest}
rf_model <- randomForest(diabetes ~ ., data = train_data,
                         ntree = 500, mtry = 3, importance = TRUE)

rf_prob_tr  <- predict(rf_model, train_data, type = "prob")[,"pos"]
rf_prob_te  <- predict(rf_model, test_data,  type = "prob")[,"pos"]
rf_class_tr <- predict(rf_model, train_data, type = "class")
rf_class_te <- predict(rf_model, test_data,  type = "class")

rf_cm_tr  <- confusionMatrix(rf_class_tr, train_data$diabetes, positive = "pos")
rf_cm_te  <- confusionMatrix(rf_class_te, test_data$diabetes,  positive = "pos")
rf_roc_tr <- roc(train_data$diabetes, rf_prob_tr, quiet = TRUE)
rf_roc_te <- roc(test_data$diabetes,  rf_prob_te, quiet = TRUE)
```

### Random Forest: Variable Importance

```{r rf-varimp, fig.height=5}
imp_df <- as.data.frame(importance(rf_model)) |>
  rownames_to_column("Variable") |>
  arrange(desc(MeanDecreaseAccuracy))

p_mda <- ggplot(imp_df, aes(x = reorder(Variable, MeanDecreaseAccuracy),
                             y = MeanDecreaseAccuracy, fill = MeanDecreaseAccuracy)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_c(option = "magma", begin = 0.3) +
  coord_flip() +
  labs(title = "Mean Decrease Accuracy", x = NULL, y = "MDA") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p_mdg <- ggplot(imp_df, aes(x = reorder(Variable, MeanDecreaseGini),
                             y = MeanDecreaseGini, fill = MeanDecreaseGini)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_c(option = "plasma", begin = 0.3) +
  coord_flip() +
  labs(title = "Mean Decrease Gini", x = NULL, y = "MDG") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

grid.arrange(p_mda, p_mdg, ncol = 2,
             top = "Random Forest Variable Importance")
```

------------------------------------------------------------------------

# Model Comparison

## Performance Summary Table

```{r comparison-table}
results <- data.frame(
  Model       = c("Logistic Regression", "LASSO", "Random Forest"),
  Train_Acc   = round(c(log_cm_tr$overall["Accuracy"],
                        las_cm_tr$overall["Accuracy"],
                        rf_cm_tr$overall["Accuracy"]), 4),
  Test_Acc    = round(c(log_cm_te$overall["Accuracy"],
                        las_cm_te$overall["Accuracy"],
                        rf_cm_te$overall["Accuracy"]), 4),
  Sensitivity = round(c(log_cm_te$byClass["Sensitivity"],
                        las_cm_te$byClass["Sensitivity"],
                        rf_cm_te$byClass["Sensitivity"]), 4),
  Specificity = round(c(log_cm_te$byClass["Specificity"],
                        las_cm_te$byClass["Specificity"],
                        rf_cm_te$byClass["Specificity"]), 4),
  F1          = round(c(log_cm_te$byClass["F1"],
                        las_cm_te$byClass["F1"],
                        rf_cm_te$byClass["F1"]), 4),
  Test_AUC    = round(c(auc(log_roc_te),
                        auc(las_roc_te),
                        auc(rf_roc_te)), 4)
)

best <- which.max(results$Test_AUC)

kbl(results, caption = "Model Performance Comparison (Test Set)",
    col.names = c("Model","Train Acc.","Test Acc.","Sensitivity",
                  "Specificity","F1 Score","Test AUC")) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE) |>
  row_spec(best, bold = TRUE, color = "white", background = palette_neg) |>
  add_footnote("Highlighted = best Test AUC", notation = "symbol")
```

## ROC Curves -- All Three Models

```{r combined-roc}
roc_df <- bind_rows(
  data.frame(fpr = 1 - log_roc_te$specificities, tpr = log_roc_te$sensitivities,
             Model = paste0("Logistic (AUC=", round(auc(log_roc_te), 3), ")")),
  data.frame(fpr = 1 - las_roc_te$specificities, tpr = las_roc_te$sensitivities,
             Model = paste0("LASSO (AUC=", round(auc(las_roc_te), 3), ")")),
  data.frame(fpr = 1 - rf_roc_te$specificities,  tpr = rf_roc_te$sensitivities,
             Model = paste0("Random Forest (AUC=", round(auc(rf_roc_te), 3), ")"))
)

ggplot(roc_df, aes(x = fpr, y = tpr, color = Model)) +
  geom_line(linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "gray60") +
  scale_color_manual(values = c(palette_neg, palette_pos, palette_neu)) +
  labs(title    = "ROC Curves -- All Models (Test Set)",
       subtitle = "Higher curve = better discrimination between diabetic and non-diabetic patients",
       x = "False Positive Rate (1 - Specificity)",
       y = "True Positive Rate (Sensitivity)",
       color = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = c(0.70, 0.18),
        legend.background = element_rect(fill = "white", color = "gray80"))
```

## Metric Bar Charts

```{r metric-bars, fig.height=5}
results |>
  select(Model, Test_Acc, Sensitivity, Specificity, Test_AUC) |>
  pivot_longer(-Model, names_to = "Metric", values_to = "Value") |>
  mutate(Metric = recode(Metric, "Test_Acc" = "Accuracy", "Test_AUC" = "AUC")) |>
  ggplot(aes(x = Model, y = Value, fill = Model)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = round(Value, 3)), vjust = -0.4, size = 3.2, fontface = "bold") +
  scale_fill_manual(values = c(palette_pos, palette_neu, palette_neg)) +
  scale_y_continuous(limits = c(0, 1.08)) +
  facet_wrap(~Metric, ncol = 4) +
  labs(title = "Test Set Performance by Model and Metric", x = NULL, y = "Score") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x  = element_text(angle = 20, hjust = 1),
        plot.title   = element_text(face = "bold"),
        strip.text   = element_text(face = "bold"))
```

## Confusion Matrices

```{r confusion-matrices, fig.height=4}
plot_cm <- function(cm_obj, title_str, color) {
  cm_df <- as.data.frame(cm_obj$table)
  cm_df$Reference  <- factor(cm_df$Reference,  levels = c("pos","neg"))
  cm_df$Prediction <- factor(cm_df$Prediction, levels = c("neg","pos"))
  ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
    geom_tile(color = "white") +
    geom_text(aes(label = Freq), size = 7, fontface = "bold") +
    scale_fill_gradient(low = "white", high = color) +
    scale_x_discrete(labels = c("pos" = "Diabetes", "neg" = "No Diabetes")) +
    scale_y_discrete(labels = c("pos" = "Diabetes", "neg" = "No Diabetes")) +
    labs(title = title_str, x = "Actual", y = "Predicted") +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(face = "bold", hjust = 0.5),
          legend.position = "none")
}

grid.arrange(
  plot_cm(log_cm_te, "Logistic Regression", palette_pos),
  plot_cm(las_cm_te, "LASSO",               palette_neu),
  plot_cm(rf_cm_te,  "Random Forest",        palette_neg),
  ncol = 3, top = "Confusion Matrices -- Test Set"
)
```

## Variable Importance Across Models

This chart normalizes importance scores within each model (so they sum to 1) to allow direct visual comparison of which predictors each model prioritizes.

```{r unified-varimp}
log_c  <- abs(coef(log_step)[-1]); log_c  <- log_c  / sum(log_c)
las_c  <- abs(as.matrix(coef(lasso_model))[-1,1])
las_c  <- las_c / sum(las_c)
rf_mda <- importance(rf_model)[,"MeanDecreaseAccuracy"]
rf_mda <- rf_mda / sum(rf_mda)

make_imp <- function(vals, model_name) {
  v <- setNames(rep(0, length(continuous_vars)), continuous_vars)
  v[names(vals)] <- vals
  data.frame(Variable = continuous_vars, Importance = v, Model = model_name)
}

all_imp <- bind_rows(
  make_imp(log_c,  "Logistic"),
  make_imp(las_c,  "LASSO"),
  make_imp(rf_mda, "Random Forest")
) |> filter(Importance > 0)

ggplot(all_imp, aes(x = reorder(Variable, Importance), y = Importance, fill = Model)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Logistic" = palette_pos, "LASSO" = palette_neu,
                               "Random Forest" = palette_neg)) +
  coord_flip() +
  labs(title    = "Normalized Variable Importance -- All Three Models",
       subtitle = "Scores scaled to sum to 1 within each model for direct comparison",
       x = NULL, y = "Normalized Importance", fill = "Model") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), legend.position = "top")
```

------------------------------------------------------------------------

# Conclusion

## Key Findings

Across all three models, **glucose** was the single most important predictor of type 2 diabetes -- consistent with its role as the primary diagnostic criterion. **BMI** and **age** ranked second and third consistently, while the **diabetes pedigree function** and **number of pregnancies** showed moderate but meaningful importance.

```{r findings-table}
data.frame(
  Rank     = 1:5,
  Variable = c("Glucose", "BMI (mass)", "Age", "Pedigree Function", "Pregnancies"),
  Reason   = c(
    "Primary clinical marker; most discriminating predictor in all three models",
    "Key driver of insulin resistance; consistently high importance",
    "Cumulative metabolic risk; strong signal across all models",
    "Captures genetic and family history component of diabetes risk",
    "Associated with gestational diabetes history and hormonal changes"
  )
) |>
  kbl(caption = "Top 5 Most Influential Variables (Consistent Across All Models)") |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE)
```

## Model Recommendation

| Model | Strengths | Best Use Case |
|------------------------|------------------------|------------------------|
| **Logistic Regression** | Interpretable; odds ratios meaningful to clinicians | When explanation is the priority |
| **LASSO** | Automatic variable selection; reduces overfitting | When a sparse, explainable model is needed |
| **Random Forest** | Highest AUC; handles non-linearity and interactions | When predictive accuracy is the priority |

Random Forest produced the best overall predictive performance. However, Logistic Regression remains valuable in clinical contexts where interpretability of individual predictors is essential for decision-making.

## Limitations

-   Dataset is limited to Pima Indian women aged 21+; results may not generalize to broader populations.
-   Class imbalance (\~35% positive) was not explicitly corrected; sensitivity could be improved with SMOTE or class weighting.
-   Imputation of missing insulin and triceps values introduces some uncertainty into the results.

------------------------------------------------------------------------

# Session Info

```{r session-info}
sessionInfo()
```
=======
---
title: "Predicting Type 2 Diabetes: A Comparative Machine Learning Analysis"
author: "Data Science Capstone Project"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Table of Contents"
    number-sections: true
    theme: cosmo
    code-fold: true
    code-tools: true
    fig-width: 9
    fig-height: 6
    fig-align: center
    embed-resources: true
    smooth-scroll: true
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

Type 2 diabetes is a major global health burden, affecting over 500 million people worldwide. Early identification of at-risk individuals allows for timely lifestyle interventions that can delay or prevent onset. This analysis uses the **Pima Indians Diabetes Dataset** — 768 health measurements from Pima Indian women aged 21 and older — to identify the most influential clinical predictors of diabetes and compare three classification models: Logistic Regression, LASSO Regression, and Random Forest.

**Research Questions:**

-   Which clinical variables are most influential in predicting type 2 diabetes?
-   Which model best balances predictive accuracy and interpretability for this clinical task?

------------------------------------------------------------------------

# Setup

```{r packages}
rm(list = ls())
set.seed(123)

required_packages <- c(
  "tidyverse", "mlbench", "caret", "pROC",
  "glmnet", "mice", "gridExtra", "reshape2",
  "randomForest", "knitr", "kableExtra", "scales", "viridis"
)

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}

# Color palette
palette_neg <- "#2E86AB"   # Blue  – No Diabetes
palette_pos <- "#A23B72"   # Purple – Diabetes
palette_neu <- "#F18F01"   # Orange – neutral / third element
```

------------------------------------------------------------------------

# Data & Preprocessing

We load the `PimaIndiansDiabetes2` dataset, which includes medically meaningful predictors and a binary outcome indicating diabetes diagnosis within five years. Several variables contain missing values (notably insulin at 49% and triceps at 30%), which are addressed using **Multiple Imputation by Chained Equations (MICE)** with predictive mean matching rather than simple deletion or zero-substitution.

```{r load-and-impute}
data("PimaIndiansDiabetes2", package = "mlbench")
diabetes_raw <- PimaIndiansDiabetes2

# Impute missing values
imp <- mice(diabetes_raw[, -9], m = 5, method = "pmm",
            seed = 123, printFlag = FALSE)
diabetes <- complete(imp, 1)
diabetes$diabetes <- diabetes_raw$diabetes

continuous_vars <- c("pregnant","glucose","pressure","triceps",
                     "insulin","mass","pedigree","age")
```

**Train / Test Split** — 70% training, 30% testing, stratified by outcome.

```{r split}
idx        <- createDataPartition(diabetes$diabetes, p = 0.70, list = FALSE)
train_data <- diabetes[idx, ]
test_data  <- diabetes[-idx, ]
```

------------------------------------------------------------------------

# Exploratory Data Analysis

## Predictor Distributions by Outcome

The density plots below reveal which variables show the clearest separation between diabetic and non-diabetic patients. Glucose, BMI, and age stand out as the most discriminating predictors.

```{r ridge-plot, fig.height=8}
diabetes |>
  select(all_of(continuous_vars), diabetes) |>
  pivot_longer(-diabetes, names_to = "Variable", values_to = "Value") |>
  mutate(Outcome = ifelse(diabetes == "neg", "No Diabetes", "Diabetes")) |>
  ggplot(aes(x = Value, fill = Outcome)) +
  geom_density(alpha = 0.65) +
  scale_fill_manual(values = c("No Diabetes" = palette_neg, "Diabetes" = palette_pos)) +
  facet_wrap(~Variable, scales = "free", ncol = 2) +
  labs(title = "Predictor Distributions by Diabetes Status",
       x = NULL, y = "Density", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "top",
        strip.text = element_text(face = "bold"))
```

## Box Plots: Top Predictors vs. Outcome

```{r boxplots, fig.height=5}
top_vars   <- c("glucose", "mass", "age", "pedigree")
top_labels <- c("Glucose (mg/dL)", "BMI (kg/m2)", "Age (years)", "Pedigree Function")

bp_list <- mapply(function(v, lab) {
  ggplot(diabetes, aes_string(x = "diabetes", y = v, fill = "diabetes")) +
    geom_boxplot(notch = TRUE, outlier.alpha = 0.4, show.legend = FALSE) +
    scale_x_discrete(labels = c("neg" = "No Diabetes", "pos" = "Diabetes")) +
    scale_fill_manual(values = c("neg" = palette_neg, "pos" = palette_pos)) +
    labs(title = lab, x = NULL, y = NULL) +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 10))
}, top_vars, top_labels, SIMPLIFY = FALSE)

grid.arrange(grobs = bp_list, ncol = 4,
             top = "Key Predictors by Diabetes Status")
```

## Correlation Heatmap

Understanding predictor correlations is important for interpreting LASSO's variable selection behavior and for detecting multicollinearity in logistic regression.

```{r heatmap, fig.height=6, fig.width=6}
cor_melt <- diabetes |>
  select(all_of(continuous_vars)) |>
  cor() |>
  melt()

ggplot(cor_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(value, 2)), size = 3, fontface = "bold") +
  scale_fill_gradient2(low = palette_neg, mid = "white", high = palette_pos,
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  labs(title = "Predictor Correlation Matrix", x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        axis.text.y = element_text(face = "bold"),
        plot.title  = element_text(face = "bold")) +
  coord_fixed()
```

## Summary Statistics

```{r summary-table}
diabetes |>
  group_by(diabetes) |>
  summarise(across(all_of(continuous_vars),
                   ~paste0(round(mean(.), 1), " (+-", round(sd(.), 1), ")")),
            .groups = "drop") |>
  mutate(diabetes = ifelse(diabetes == "neg", "No Diabetes", "Diabetes")) |>
  pivot_longer(-diabetes, names_to = "Variable") |>
  pivot_wider(names_from = diabetes, values_from = value) |>
  kbl(caption = "Mean (+-SD) by Diabetes Status",
      col.names = c("Variable", "No Diabetes", "Diabetes")) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

------------------------------------------------------------------------

# Model Building

## Logistic Regression

Logistic regression serves as our interpretable baseline. We apply bidirectional stepwise selection (AIC) to retain only the most informative predictors.

```{r logistic}
log_full <- glm(diabetes ~ ., data = train_data, family = binomial)
log_step <- step(log_full, direction = "both", trace = 0)

# Predictions
log_prob_tr  <- predict(log_step, train_data, type = "response")
log_prob_te  <- predict(log_step, test_data,  type = "response")
log_class_tr <- factor(ifelse(log_prob_tr > 0.5, "pos", "neg"), levels = c("neg","pos"))
log_class_te <- factor(ifelse(log_prob_te > 0.5, "pos", "neg"), levels = c("neg","pos"))

log_cm_tr  <- confusionMatrix(log_class_tr, train_data$diabetes, positive = "pos")
log_cm_te  <- confusionMatrix(log_class_te, test_data$diabetes,  positive = "pos")
log_roc_tr <- roc(train_data$diabetes, log_prob_tr, quiet = TRUE)
log_roc_te <- roc(test_data$diabetes,  log_prob_te, quiet = TRUE)
```

### Odds Ratios

Variables with OR \> 1 and a confidence interval that does not cross 1 are statistically significant risk factors.

```{r odds-ratios}
or_df <- data.frame(
  Variable = names(coef(log_step)),
  OR       = exp(coef(log_step)),
  CI_Lo    = exp(confint.default(log_step)[,1]),
  CI_Hi    = exp(confint.default(log_step)[,2])
) |> filter(Variable != "(Intercept)") |>
  mutate(across(where(is.numeric), ~round(., 3)))

kbl(or_df, caption = "Logistic Regression: Odds Ratios with 95% CI",
    col.names = c("Variable","OR","Lower CI","Upper CI")) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) |>
  row_spec(which(or_df$OR > 1 & or_df$CI_Lo > 1), bold = TRUE,
           color = "white", background = palette_pos)

ggplot(or_df, aes(x = reorder(Variable, OR), y = OR,
                  ymin = CI_Lo, ymax = CI_Hi, color = OR > 1)) +
  geom_pointrange(size = 0.9, linewidth = 1) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("FALSE" = palette_neg, "TRUE" = palette_pos),
                     labels = c("No/Neutral Effect", "Increased Risk"), name = NULL) +
  coord_flip() +
  labs(title    = "Logistic Regression: Odds Ratios with 95% CI",
       subtitle = "Purple = statistically significant risk factor (OR > 1, CI excludes 1)",
       x = NULL, y = "Odds Ratio") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), legend.position = "top")
```

------------------------------------------------------------------------

## LASSO Regression

LASSO applies an L1 penalty that shrinks weaker coefficients to zero, producing an automatically selected subset of predictors. The optimal penalty (lambda) is chosen via 10-fold cross-validation.

```{r lasso}
x_tr <- model.matrix(diabetes ~ ., train_data)[,-1]
y_tr <- ifelse(train_data$diabetes == "pos", 1, 0)
x_te <- model.matrix(diabetes ~ ., test_data)[,-1]

cv_lasso    <- cv.glmnet(x_tr, y_tr, alpha = 1, family = "binomial", nfolds = 10)
lasso_model <- glmnet(x_tr, y_tr, alpha = 1, family = "binomial",
                      lambda = cv_lasso$lambda.min)

# Predictions
las_prob_tr  <- predict(lasso_model, x_tr, type = "response")[,1]
las_prob_te  <- predict(lasso_model, x_te, type = "response")[,1]
las_class_tr <- factor(ifelse(las_prob_tr > 0.5, "pos", "neg"), levels = c("neg","pos"))
las_class_te <- factor(ifelse(las_prob_te > 0.5, "pos", "neg"), levels = c("neg","pos"))

las_cm_tr  <- confusionMatrix(las_class_tr, train_data$diabetes, positive = "pos")
las_cm_te  <- confusionMatrix(las_class_te, test_data$diabetes,  positive = "pos")
las_roc_tr <- roc(train_data$diabetes, las_prob_tr, quiet = TRUE)
las_roc_te <- roc(test_data$diabetes,  las_prob_te, quiet = TRUE)
```

### LASSO: Selected Coefficients

```{r lasso-coef-plot}
las_coef_df <- data.frame(
  Variable = rownames(as.matrix(coef(lasso_model))),
  Coef     = as.matrix(coef(lasso_model))[,1]
) |> filter(Variable != "(Intercept)", Coef != 0) |>
  arrange(desc(abs(Coef)))

ggplot(las_coef_df, aes(x = reorder(Variable, abs(Coef)), y = Coef, fill = Coef > 0)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("FALSE" = palette_neg, "TRUE" = palette_pos)) +
  coord_flip() +
  labs(title    = "LASSO: Non-Zero Coefficients After Penalization",
       subtitle = "Purple = increases risk; Blue = decreases risk; absent = shrunk to zero",
       x = NULL, y = "Coefficient") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

------------------------------------------------------------------------

## Random Forest

Random Forest is an ensemble of 500 decision trees that reduces variance through bagging and random feature subsampling. It naturally captures non-linear relationships and feature interactions that linear models may miss.

```{r random-forest}
rf_model <- randomForest(diabetes ~ ., data = train_data,
                         ntree = 500, mtry = 3, importance = TRUE)

rf_prob_tr  <- predict(rf_model, train_data, type = "prob")[,"pos"]
rf_prob_te  <- predict(rf_model, test_data,  type = "prob")[,"pos"]
rf_class_tr <- predict(rf_model, train_data, type = "class")
rf_class_te <- predict(rf_model, test_data,  type = "class")

rf_cm_tr  <- confusionMatrix(rf_class_tr, train_data$diabetes, positive = "pos")
rf_cm_te  <- confusionMatrix(rf_class_te, test_data$diabetes,  positive = "pos")
rf_roc_tr <- roc(train_data$diabetes, rf_prob_tr, quiet = TRUE)
rf_roc_te <- roc(test_data$diabetes,  rf_prob_te, quiet = TRUE)
```

### Random Forest: Variable Importance

```{r rf-varimp, fig.height=5}
imp_df <- as.data.frame(importance(rf_model)) |>
  rownames_to_column("Variable") |>
  arrange(desc(MeanDecreaseAccuracy))

p_mda <- ggplot(imp_df, aes(x = reorder(Variable, MeanDecreaseAccuracy),
                             y = MeanDecreaseAccuracy, fill = MeanDecreaseAccuracy)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_c(option = "magma", begin = 0.3) +
  coord_flip() +
  labs(title = "Mean Decrease Accuracy", x = NULL, y = "MDA") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p_mdg <- ggplot(imp_df, aes(x = reorder(Variable, MeanDecreaseGini),
                             y = MeanDecreaseGini, fill = MeanDecreaseGini)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_c(option = "plasma", begin = 0.3) +
  coord_flip() +
  labs(title = "Mean Decrease Gini", x = NULL, y = "MDG") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

grid.arrange(p_mda, p_mdg, ncol = 2,
             top = "Random Forest Variable Importance")
```

------------------------------------------------------------------------

# Model Comparison

## Performance Summary Table

```{r comparison-table}
results <- data.frame(
  Model       = c("Logistic Regression", "LASSO", "Random Forest"),
  Train_Acc   = round(c(log_cm_tr$overall["Accuracy"],
                        las_cm_tr$overall["Accuracy"],
                        rf_cm_tr$overall["Accuracy"]), 4),
  Test_Acc    = round(c(log_cm_te$overall["Accuracy"],
                        las_cm_te$overall["Accuracy"],
                        rf_cm_te$overall["Accuracy"]), 4),
  Sensitivity = round(c(log_cm_te$byClass["Sensitivity"],
                        las_cm_te$byClass["Sensitivity"],
                        rf_cm_te$byClass["Sensitivity"]), 4),
  Specificity = round(c(log_cm_te$byClass["Specificity"],
                        las_cm_te$byClass["Specificity"],
                        rf_cm_te$byClass["Specificity"]), 4),
  F1          = round(c(log_cm_te$byClass["F1"],
                        las_cm_te$byClass["F1"],
                        rf_cm_te$byClass["F1"]), 4),
  Test_AUC    = round(c(auc(log_roc_te),
                        auc(las_roc_te),
                        auc(rf_roc_te)), 4)
)

best <- which.max(results$Test_AUC)

kbl(results, caption = "Model Performance Comparison (Test Set)",
    col.names = c("Model","Train Acc.","Test Acc.","Sensitivity",
                  "Specificity","F1 Score","Test AUC")) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE) |>
  row_spec(best, bold = TRUE, color = "white", background = palette_neg) |>
  add_footnote("Highlighted = best Test AUC", notation = "symbol")
```

## ROC Curves -- All Three Models

```{r combined-roc}
roc_df <- bind_rows(
  data.frame(fpr = 1 - log_roc_te$specificities, tpr = log_roc_te$sensitivities,
             Model = paste0("Logistic (AUC=", round(auc(log_roc_te), 3), ")")),
  data.frame(fpr = 1 - las_roc_te$specificities, tpr = las_roc_te$sensitivities,
             Model = paste0("LASSO (AUC=", round(auc(las_roc_te), 3), ")")),
  data.frame(fpr = 1 - rf_roc_te$specificities,  tpr = rf_roc_te$sensitivities,
             Model = paste0("Random Forest (AUC=", round(auc(rf_roc_te), 3), ")"))
)

ggplot(roc_df, aes(x = fpr, y = tpr, color = Model)) +
  geom_line(linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "gray60") +
  scale_color_manual(values = c(palette_neg, palette_pos, palette_neu)) +
  labs(title    = "ROC Curves -- All Models (Test Set)",
       subtitle = "Higher curve = better discrimination between diabetic and non-diabetic patients",
       x = "False Positive Rate (1 - Specificity)",
       y = "True Positive Rate (Sensitivity)",
       color = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = c(0.70, 0.18),
        legend.background = element_rect(fill = "white", color = "gray80"))
```

## Metric Bar Charts

```{r metric-bars, fig.height=5}
results |>
  select(Model, Test_Acc, Sensitivity, Specificity, Test_AUC) |>
  pivot_longer(-Model, names_to = "Metric", values_to = "Value") |>
  mutate(Metric = recode(Metric, "Test_Acc" = "Accuracy", "Test_AUC" = "AUC")) |>
  ggplot(aes(x = Model, y = Value, fill = Model)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = round(Value, 3)), vjust = -0.4, size = 3.2, fontface = "bold") +
  scale_fill_manual(values = c(palette_pos, palette_neu, palette_neg)) +
  scale_y_continuous(limits = c(0, 1.08)) +
  facet_wrap(~Metric, ncol = 4) +
  labs(title = "Test Set Performance by Model and Metric", x = NULL, y = "Score") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x  = element_text(angle = 20, hjust = 1),
        plot.title   = element_text(face = "bold"),
        strip.text   = element_text(face = "bold"))
```

## Confusion Matrices

```{r confusion-matrices, fig.height=4}
plot_cm <- function(cm_obj, title_str, color) {
  cm_df <- as.data.frame(cm_obj$table)
  cm_df$Reference  <- factor(cm_df$Reference,  levels = c("pos","neg"))
  cm_df$Prediction <- factor(cm_df$Prediction, levels = c("neg","pos"))
  ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
    geom_tile(color = "white") +
    geom_text(aes(label = Freq), size = 7, fontface = "bold") +
    scale_fill_gradient(low = "white", high = color) +
    scale_x_discrete(labels = c("pos" = "Diabetes", "neg" = "No Diabetes")) +
    scale_y_discrete(labels = c("pos" = "Diabetes", "neg" = "No Diabetes")) +
    labs(title = title_str, x = "Actual", y = "Predicted") +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(face = "bold", hjust = 0.5),
          legend.position = "none")
}

grid.arrange(
  plot_cm(log_cm_te, "Logistic Regression", palette_pos),
  plot_cm(las_cm_te, "LASSO",               palette_neu),
  plot_cm(rf_cm_te,  "Random Forest",        palette_neg),
  ncol = 3, top = "Confusion Matrices -- Test Set"
)
```

## Variable Importance Across Models

This chart normalizes importance scores within each model (so they sum to 1) to allow direct visual comparison of which predictors each model prioritizes.

```{r unified-varimp}
log_c  <- abs(coef(log_step)[-1]); log_c  <- log_c  / sum(log_c)
las_c  <- abs(as.matrix(coef(lasso_model))[-1,1])
las_c  <- las_c / sum(las_c)
rf_mda <- importance(rf_model)[,"MeanDecreaseAccuracy"]
rf_mda <- rf_mda / sum(rf_mda)

make_imp <- function(vals, model_name) {
  v <- setNames(rep(0, length(continuous_vars)), continuous_vars)
  v[names(vals)] <- vals
  data.frame(Variable = continuous_vars, Importance = v, Model = model_name)
}

all_imp <- bind_rows(
  make_imp(log_c,  "Logistic"),
  make_imp(las_c,  "LASSO"),
  make_imp(rf_mda, "Random Forest")
) |> filter(Importance > 0)

ggplot(all_imp, aes(x = reorder(Variable, Importance), y = Importance, fill = Model)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Logistic" = palette_pos, "LASSO" = palette_neu,
                               "Random Forest" = palette_neg)) +
  coord_flip() +
  labs(title    = "Normalized Variable Importance -- All Three Models",
       subtitle = "Scores scaled to sum to 1 within each model for direct comparison",
       x = NULL, y = "Normalized Importance", fill = "Model") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), legend.position = "top")
```

------------------------------------------------------------------------

# Conclusion

## Key Findings

Across all three models, **glucose** was the single most important predictor of type 2 diabetes -- consistent with its role as the primary diagnostic criterion. **BMI** and **age** ranked second and third consistently, while the **diabetes pedigree function** and **number of pregnancies** showed moderate but meaningful importance.

```{r findings-table}
data.frame(
  Rank     = 1:5,
  Variable = c("Glucose", "BMI (mass)", "Age", "Pedigree Function", "Pregnancies"),
  Reason   = c(
    "Primary clinical marker; most discriminating predictor in all three models",
    "Key driver of insulin resistance; consistently high importance",
    "Cumulative metabolic risk; strong signal across all models",
    "Captures genetic and family history component of diabetes risk",
    "Associated with gestational diabetes history and hormonal changes"
  )
) |>
  kbl(caption = "Top 5 Most Influential Variables (Consistent Across All Models)") |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE)
```

## Model Recommendation

| Model | Strengths | Best Use Case |
|------------------------|------------------------|------------------------|
| **Logistic Regression** | Interpretable; odds ratios meaningful to clinicians | When explanation is the priority |
| **LASSO** | Automatic variable selection; reduces overfitting | When a sparse, explainable model is needed |
| **Random Forest** | Highest AUC; handles non-linearity and interactions | When predictive accuracy is the priority |

Random Forest produced the best overall predictive performance. However, Logistic Regression remains valuable in clinical contexts where interpretability of individual predictors is essential for decision-making.

## Limitations

-   Dataset is limited to Pima Indian women aged 21+; results may not generalize to broader populations.
-   Class imbalance (\~35% positive) was not explicitly corrected; sensitivity could be improved with SMOTE or class weighting.
-   Imputation of missing insulin and triceps values introduces some uncertainty into the results.

------------------------------------------------------------------------

# Session Info

```{r session-info}
sessionInfo()
```
>>>>>>> a3c158bb79071ec11925bc7e63e47b6ba82e5726
