---
title: "Predicting Type 2 Diabetes with Machine Learning"
subtitle: "A Comparative Analysis of Logistic Regression, LASSO, and Random Forest"
author: "Data Science Capstone Project"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Table of Contents"
    number-sections: true
    theme: cosmo
    code-fold: true
    code-tools: true
    fig-width: 9
    fig-height: 6
    fig-align: center
    embed-resources: true
    smooth-scroll: true
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

Type 2 diabetes is a major global health burden, affecting over 500 million people worldwide. Early identification of at-risk individuals allows for timely lifestyle interventions that can delay or prevent onset. This analysis uses the **Pima Indians Diabetes Dataset** — 768 health measurements from Pima Indian women aged 21 and older — to identify the most influential clinical predictors of diabetes and compare three classification models: Logistic Regression, LASSO Regression, and Random Forest.

**Research Questions:**

-   Which clinical variables are most influential in predicting type 2 diabetes?
-   Which model best balances predictive accuracy and interpretability for this clinical task?

------------------------------------------------------------------------

# Setup

```{r packages}
rm(list = ls())
set.seed(123)

required_packages <- c(
  "tidyverse", "mlbench", "caret", "pROC",
  "glmnet", "mice", "gridExtra", "reshape2",
  "randomForest", "knitr", "kableExtra", "scales", "viridis"
)

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}

# Consistent color palette used throughout all visuals
palette_neg <- "#2E86AB"   # Blue   – No Diabetes
palette_pos <- "#A23B72"   # Purple – Diabetes
palette_neu <- "#F18F01"   # Orange – neutral / third model
```

------------------------------------------------------------------------

# Data & Preprocessing

We load the `PimaIndiansDiabetes2` dataset, which includes medically meaningful predictors and a binary outcome indicating diabetes diagnosis within five years. Several variables contain missing values — notably insulin (49% missing) and triceps skinfold (30% missing). Rather than dropping rows or substituting zeros (which would introduce systematic bias), we address this using **Multiple Imputation by Chained Equations (MICE)** with predictive mean matching. This preserves the natural variability of the data while producing a complete dataset suitable for modeling.

```{r load-and-impute}
data("PimaIndiansDiabetes2", package = "mlbench")
diabetes_raw <- PimaIndiansDiabetes2

# Impute missing values with MICE (predictive mean matching, 5 imputations)
imp <- mice(diabetes_raw[, -9], m = 5, method = "pmm",
            seed = 123, printFlag = FALSE)
diabetes <- complete(imp, 1)
diabetes$diabetes <- diabetes_raw$diabetes

continuous_vars <- c("pregnant", "glucose", "pressure", "triceps",
                     "insulin", "mass", "pedigree", "age")
```

**Train / Test Split** — 70% training, 30% testing, stratified by outcome to preserve the natural class ratio in both sets.

```{r split}
idx        <- createDataPartition(diabetes$diabetes, p = 0.70, list = FALSE)
train_data <- diabetes[idx, ]
test_data  <- diabetes[-idx, ]
```

------------------------------------------------------------------------

# Exploratory Data Analysis

## Predictor Distributions by Outcome

The density plots below reveal the separation between diabetic and non-diabetic patients across all clinical variables. **Glucose** shows the sharpest separation, followed by **BMI** and **age** — suggesting these will be the most discriminating predictors across all models.

```{r ridge-plot, fig.height=8}
diabetes |>
  select(all_of(continuous_vars), diabetes) |>
  pivot_longer(-diabetes, names_to = "Variable", values_to = "Value") |>
  mutate(Outcome = ifelse(diabetes == "neg", "No Diabetes", "Diabetes")) |>
  ggplot(aes(x = Value, fill = Outcome)) +
  geom_density(alpha = 0.65) +
  scale_fill_manual(values = c("No Diabetes" = palette_neg, "Diabetes" = palette_pos)) +
  facet_wrap(~Variable, scales = "free", ncol = 2) +
  labs(title    = "Predictor Distributions by Diabetes Status",
       subtitle = "Clear separation in glucose, BMI, and age signals strong predictive potential",
       x = NULL, y = "Density", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(color = "gray40"),
        legend.position = "top")
```

## Box Plots: Key Predictors vs. Outcome

These box plots focus on the four variables expected to drive model performance. The notches indicate 95% confidence intervals around the median — non-overlapping notches indicate statistically meaningful differences between groups.

```{r boxplots, fig.height=5}
top_vars   <- c("glucose", "mass", "age", "pedigree")
top_labels <- c("Glucose (mg/dL)", "BMI (kg/m²)", "Age (years)", "Diabetes Pedigree Function")

bp_list <- mapply(function(v, lab) {
  ggplot(diabetes, aes(x = diabetes, y = .data[[v]], fill = diabetes)) +
    geom_boxplot(notch = TRUE, outlier.alpha = 0.4, show.legend = FALSE) +
    scale_x_discrete(labels = c("neg" = "No Diabetes", "pos" = "Diabetes")) +
    scale_fill_manual(values = c("neg" = palette_neg, "pos" = palette_pos)) +
    labs(title = lab, x = NULL, y = NULL) +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 10))
}, top_vars, top_labels, SIMPLIFY = FALSE)

grid.arrange(grobs = bp_list, ncol = 4,
             top = "Key Predictors by Diabetes Status")
```

## Correlation Heatmap

Understanding predictor correlations informs how we interpret LASSO's variable selection and flags potential multicollinearity in logistic regression. The strongest correlations appear between age and number of pregnancies, and between BMI and triceps skinfold — both clinically expected.

```{r heatmap, fig.height=6, fig.width=6}
cor_melt <- diabetes |>
  select(all_of(continuous_vars)) |>
  cor() |>
  melt()

ggplot(cor_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(value, 2)), size = 3, fontface = "bold") +
  scale_fill_gradient2(low = palette_neg, mid = "white", high = palette_pos,
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  labs(title = "Predictor Correlation Matrix", x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        axis.text.y = element_text(face = "bold"),
        plot.title  = element_text(face = "bold")) +
  coord_fixed()
```

## Summary Statistics

```{r summary-table}
diabetes |>
  group_by(diabetes) |>
  summarise(across(all_of(continuous_vars),
                   ~paste0(round(mean(.), 1), " (\u00B1", round(sd(.), 1), ")")),
            .groups = "drop") |>
  mutate(diabetes = ifelse(diabetes == "neg", "No Diabetes", "Diabetes")) |>
  pivot_longer(-diabetes, names_to = "Variable") |>
  pivot_wider(names_from = diabetes, values_from = value) |>
  kbl(caption = "Mean (\u00B1 SD) by Diabetes Status") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

------------------------------------------------------------------------

## Logistic Regression

Logistic regression serves as our interpretable baseline. We apply **bidirectional stepwise selection** (AIC criterion) to automatically retain only the most informative predictors and reduce overfitting risk.

```{r logistic}
log_full <- glm(diabetes ~ ., data = train_data, family = binomial)
log_step <- step(log_full, direction = "both", trace = 0)

# Predictions on train and test
log_prob_tr  <- predict(log_step, train_data, type = "response")
log_prob_te  <- predict(log_step, test_data,  type = "response")
log_class_tr <- factor(ifelse(log_prob_tr > 0.5, "pos", "neg"), levels = c("neg", "pos"))
log_class_te <- factor(ifelse(log_prob_te > 0.5, "pos", "neg"), levels = c("neg", "pos"))

log_cm_tr  <- confusionMatrix(log_class_tr, train_data$diabetes, positive = "pos")
log_cm_te  <- confusionMatrix(log_class_te, test_data$diabetes,  positive = "pos")
log_roc_tr <- roc(train_data$diabetes, log_prob_tr, quiet = TRUE)
log_roc_te <- roc(test_data$diabetes,  log_prob_te, quiet = TRUE)
```

## LASSO Regression

LASSO applies an **L1 penalty** that progressively shrinks weaker coefficients to exactly zero, producing automatic variable selection. The optimal penalty strength (lambda) is chosen via **10-fold cross-validation** to minimize binomial deviance.

```{r lasso}
x_tr <- model.matrix(diabetes ~ ., train_data)[, -1]
y_tr <- ifelse(train_data$diabetes == "pos", 1, 0)
x_te <- model.matrix(diabetes ~ ., test_data)[, -1]

cv_lasso    <- cv.glmnet(x_tr, y_tr, alpha = 1, family = "binomial", nfolds = 10)
lasso_model <- glmnet(x_tr, y_tr, alpha = 1, family = "binomial",
                      lambda = cv_lasso$lambda.min)

# Predictions
las_prob_tr  <- predict(lasso_model, x_tr, type = "response")[, 1]
las_prob_te  <- predict(lasso_model, x_te, type = "response")[, 1]
las_class_tr <- factor(ifelse(las_prob_tr > 0.5, "pos", "neg"), levels = c("neg", "pos"))
las_class_te <- factor(ifelse(las_prob_te > 0.5, "pos", "neg"), levels = c("neg", "pos"))

las_cm_tr  <- confusionMatrix(las_class_tr, train_data$diabetes, positive = "pos")
las_cm_te  <- confusionMatrix(las_class_te, test_data$diabetes,  positive = "pos")
las_roc_tr <- roc(train_data$diabetes, las_prob_tr, quiet = TRUE)
las_roc_te <- roc(test_data$diabetes,  las_prob_te, quiet = TRUE)
```

## Random Forest

Random Forest is an **ensemble of 500 decision trees** that reduces variance through bagging and random feature subsampling at each split. It naturally captures non-linear relationships and feature interactions that linear models may miss. We set `mtry = 3` (the square root of 8 predictors) following standard practice.

```{r random-forest}
rf_model <- randomForest(diabetes ~ ., data = train_data,
                         ntree = 500, mtry = 3, importance = TRUE)

rf_prob_tr  <- predict(rf_model, train_data, type = "prob")[, "pos"]
rf_prob_te  <- predict(rf_model, test_data,  type = "prob")[, "pos"]
rf_class_tr <- predict(rf_model, train_data, type = "class")
rf_class_te <- predict(rf_model, test_data,  type = "class")

rf_cm_tr  <- confusionMatrix(rf_class_tr, train_data$diabetes, positive = "pos")
rf_cm_te  <- confusionMatrix(rf_class_te, test_data$diabetes,  positive = "pos")
rf_roc_tr <- roc(train_data$diabetes, rf_prob_tr, quiet = TRUE)
rf_roc_te <- roc(test_data$diabetes,  rf_prob_te, quiet = TRUE)
```

### Random Forest: Out-of-Bag Error

The OOB error curve shows how model performance stabilizes as more trees are added. Convergence before 500 trees indicates the ensemble is well-trained and not under-fit.

```{r rf-oob-plot, fig.height=4}
oob_df <- data.frame(
  Trees = seq_len(nrow(rf_model$err.rate)),
  OOB   = rf_model$err.rate[, "OOB"],
  neg   = rf_model$err.rate[, "neg"],
  pos   = rf_model$err.rate[, "pos"]
) |>
  pivot_longer(-Trees, names_to = "Class", values_to = "Error") |>
  mutate(Class = recode(Class, "OOB" = "Overall OOB",
                        "neg" = "No Diabetes", "pos" = "Diabetes"))

ggplot(oob_df, aes(x = Trees, y = Error, color = Class)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c("Overall OOB" = "black",
                                "No Diabetes" = palette_neg,
                                "Diabetes"    = palette_pos)) +
  labs(title    = "Random Forest: Out-of-Bag Error Rate by Number of Trees",
       subtitle = "Error stabilizes well before 500 trees",
       x = "Number of Trees", y = "Error Rate", color = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(color = "gray40"),
        legend.position = "top")
```

### Random Forest: Variable Importance

Two complementary importance metrics are shown. **Mean Decrease Accuracy (MDA)** measures how much each variable's removal degrades prediction accuracy. **Mean Decrease Gini (MDG)** measures each variable's contribution to node purity during splitting. Agreement between both metrics strengthens confidence in the rankings.

```{r rf-varimp, fig.height=5}
imp_df <- as.data.frame(importance(rf_model)) |>
  rownames_to_column("Variable") |>
  arrange(desc(MeanDecreaseAccuracy))

p_mda <- ggplot(imp_df, aes(x = reorder(Variable, MeanDecreaseAccuracy),
                             y = MeanDecreaseAccuracy,
                             fill = MeanDecreaseAccuracy)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_c(option = "magma", begin = 0.3) +
  coord_flip() +
  labs(title = "Mean Decrease Accuracy", x = NULL, y = "MDA") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p_mdg <- ggplot(imp_df, aes(x = reorder(Variable, MeanDecreaseGini),
                             y = MeanDecreaseGini,
                             fill = MeanDecreaseGini)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_c(option = "plasma", begin = 0.3) +
  coord_flip() +
  labs(title = "Mean Decrease Gini", x = NULL, y = "MDG") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

grid.arrange(p_mda, p_mdg, ncol = 2,
             top = "Random Forest Variable Importance")
```

------------------------------------------------------------------------

# Model Comparison

## Performance Summary Table

The table below compares all three models on the held-out test set. The best-performing model by Test AUC is highlighted.

```{r comparison-table}
results <- data.frame(
  Model       = c("Logistic Regression", "LASSO", "Random Forest"),
  Train_Acc   = round(c(log_cm_tr$overall["Accuracy"],
                        las_cm_tr$overall["Accuracy"],
                        rf_cm_tr$overall["Accuracy"]), 4),
  Test_Acc    = round(c(log_cm_te$overall["Accuracy"],
                        las_cm_te$overall["Accuracy"],
                        rf_cm_te$overall["Accuracy"]), 4),
  Sensitivity = round(c(log_cm_te$byClass["Sensitivity"],
                        las_cm_te$byClass["Sensitivity"],
                        rf_cm_te$byClass["Sensitivity"]), 4),
  Specificity = round(c(log_cm_te$byClass["Specificity"],
                        las_cm_te$byClass["Specificity"],
                        rf_cm_te$byClass["Specificity"]), 4),
  F1          = round(c(log_cm_te$byClass["F1"],
                        las_cm_te$byClass["F1"],
                        rf_cm_te$byClass["F1"]), 4),
  Test_AUC    = round(c(auc(log_roc_te),
                        auc(las_roc_te),
                        auc(rf_roc_te)), 4)
)

best <- which.max(results$Test_AUC)

kbl(results, caption = "Model Performance Comparison (Test Set)",
    col.names = c("Model", "Train Acc.", "Test Acc.", "Sensitivity",
                  "Specificity", "F1 Score", "Test AUC")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) |>
  row_spec(best, bold = TRUE, color = "white", background = palette_neg) |>
  add_footnote("Highlighted row = best Test AUC", notation = "symbol")
```

## ROC Curves — All Three Models

The ROC curve plots sensitivity against the false positive rate across all classification thresholds. A curve closer to the top-left corner indicates a stronger model. The diagonal dashed line represents random chance (AUC = 0.50).

```{r combined-roc, fig.height=5}
roc_df <- bind_rows(
  data.frame(fpr = 1 - log_roc_te$specificities, tpr = log_roc_te$sensitivities,
             Model = paste0("Logistic (AUC = ", round(auc(log_roc_te), 3), ")")),
  data.frame(fpr = 1 - las_roc_te$specificities, tpr = las_roc_te$sensitivities,
             Model = paste0("LASSO (AUC = ", round(auc(las_roc_te), 3), ")")),
  data.frame(fpr = 1 - rf_roc_te$specificities,  tpr = rf_roc_te$sensitivities,
             Model = paste0("Random Forest (AUC = ", round(auc(rf_roc_te), 3), ")"))
)

ggplot(roc_df, aes(x = fpr, y = tpr, color = Model)) +
  geom_line(linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "gray60") +
  scale_color_manual(values = c(palette_pos, palette_neu, palette_neg)) +
  labs(title    = "ROC Curves — All Models (Test Set)",
       subtitle = "Higher curve = better discrimination between diabetic and non-diabetic patients",
       x = "False Positive Rate (1 – Specificity)",
       y = "True Positive Rate (Sensitivity)",
       color = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(color = "gray40"),
        legend.position = c(0.68, 0.18),
        legend.background = element_rect(fill = "white", color = "gray80"))
```

## Metric Bar Charts

```{r metric-bars, fig.height=5}
results |>
  select(Model, Test_Acc, Sensitivity, Specificity, Test_AUC) |>
  pivot_longer(-Model, names_to = "Metric", values_to = "Value") |>
  mutate(Metric = recode(Metric,
                         "Test_Acc" = "Accuracy",
                         "Test_AUC" = "AUC")) |>
  ggplot(aes(x = Model, y = Value, fill = Model)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = round(Value, 3)), vjust = -0.4, size = 3.2, fontface = "bold") +
  scale_fill_manual(values = c(palette_pos, palette_neu, palette_neg)) +
  scale_y_continuous(limits = c(0, 1.08)) +
  facet_wrap(~Metric, ncol = 4) +
  labs(title = "Test Set Performance by Model and Metric", x = NULL, y = "Score") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1),
        plot.title  = element_text(face = "bold"),
        strip.text  = element_text(face = "bold"))
```

## Normalized Variable Importance — All Three Models

To allow direct comparison across models, importance scores are **standardized so they sum to 1** within each model. Logistic regression and LASSO are refit on scaled predictors so their coefficients are on the same unit-free scale. This ensures differences in bar height reflect genuine differences in predictor importance, not differences in variable units.

```{r unified-varimp}
# Refit Logistic and LASSO on standardized predictors for fair comparison
# (original models remain intact for odds ratios and predictions)
train_scaled <- train_data
train_scaled[continuous_vars] <- scale(train_data[continuous_vars])

log_scaled <- glm(formula(log_step), data = train_scaled, family = binomial)

x_tr_sc    <- model.matrix(diabetes ~ ., train_scaled)[, -1]
las_scaled <- glmnet(x_tr_sc, y_tr, alpha = 1, family = "binomial",
                     lambda = cv_lasso$lambda.min)

# Normalize so each model's scores sum to 1
log_c  <- abs(coef(log_scaled)[-1])
log_c  <- log_c / sum(log_c)

las_c  <- abs(as.matrix(coef(las_scaled))[-1, 1])
las_c  <- las_c[las_c != 0]
las_c  <- las_c / sum(las_c)

rf_mda <- importance(rf_model)[, "MeanDecreaseAccuracy"]
rf_mda <- rf_mda / sum(rf_mda)

make_imp <- function(vals, model_name) {
  v <- setNames(rep(0, length(continuous_vars)), continuous_vars)
  v[intersect(names(vals), continuous_vars)] <- vals[intersect(names(vals), continuous_vars)]
  data.frame(Variable = continuous_vars, Importance = v, Model = model_name)
}

all_imp <- bind_rows(
  make_imp(log_c,  "Logistic"),
  make_imp(las_c,  "LASSO"),
  make_imp(rf_mda, "Random Forest")
) |> filter(Importance > 0)

ggplot(all_imp, aes(x = reorder(Variable, Importance), y = Importance, fill = Model)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Logistic"      = palette_pos,
                               "LASSO"         = palette_neu,
                               "Random Forest" = palette_neg)) +
  coord_flip() +
  labs(title    = "Normalized Variable Importance — All Three Models",
       subtitle = "Logistic & LASSO refit on standardized predictors for unit-free comparison",
       x = NULL, y = "Normalized Importance (sums to 1 per model)", fill = "Model") +
  theme_minimal(base_size = 13) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(color = "gray40"),
        legend.position = "top")
```

------------------------------------------------------------------------

# Conclusion

## Key Findings

Across all three models, **glucose** was the single most important predictor of type 2 diabetes — consistent with its central role as the primary clinical diagnostic criterion. **BMI** and **age** ranked second and third consistently, while the **diabetes pedigree function** (a measure of genetic risk) and **number of pregnancies** showed moderate but meaningful predictive value.

```{r findings-table}
data.frame(
  Rank      = 1:5,
  Variable  = c("Glucose", "BMI (mass)", "Age", "Diabetes Pedigree Function", "Pregnancies"),
  Rationale = c(
    "Primary clinical marker; clearest separation between groups",
    "Obesity is a leading modifiable risk factor for T2D",
    "Risk accumulates with age; older patients show higher prevalence",
    "Captures heritable family history of diabetes",
    "Associated with gestational diabetes and hormonal changes"
  )
) |>
  kbl(caption = "Top 5 Most Influential Variables (Consistent Across All Models)") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

## Model Recommendation

| Model | Test AUC | Strengths | Best Use Case |
|------------------|------------------|------------------|------------------|
| **LASSO** | 0.845 | Highest AUC; automatic variable selection; sparse and interpretable | Best overall performer; ideal when accuracy and simplicity both matter |
| **Logistic Regression** | 0.842 | Nearly identical AUC; odds ratios clinically communicable | When individual predictor effects must be explained to patients or regulators |
| **Random Forest** | 0.827 | Best threshold-based sensitivity; captures non-linear structure | Larger datasets where non-linearity can be learned reliably |
